The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-02-10 src/ms.tex
New features as itemized list

% Links to documentation: subpackages
\newcommand{\astropysubpkg}[1]{\href{http://docs.astropy.org/en/stable/#1/index.html}{\texttt{astropy.#1}}\xspace}
\newcommand{\astropyiosubpkg}[1]{\href{http://docs.astropy.org/en/stable/io/#1/index.html}{\texttt{astropy.io.#1}}\xspace}
\newcommand{\astropywcsaxes}{\href{http://docs.astropy.org/en/stable/visualization/wcsaxes/index.html}{\texttt{astropy.visualization.wcsaxes}}\xspace}

% Links to documentation: classes
\newcommand{\astropyskycoord}{\href{http://docs.astropy.org/en/stable/api/astropy.coordinates.SkyCoord.html}{\texttt{SkyCoord}}\xspace}
\newcommand{\astropyQuantity}{\href{http://docs.astropy.org/en/stable/api/astropy.units.Quantity.html}{\texttt{Quantity}}\xspace}
\newcommand{\astropyTime}{\href{http://docs.astropy.org/en/stable/api/astropy.time.Time.html}{\texttt{Time}}\xspace}
\newcommand{\astropyTable}{\href{http://docs.astropy.org/en/stable/api/astropy.table.Table.html}{\texttt{Table}}\xspace}

@@

Summarize major pieces of development since the last paper (since v2.0). A few
ideas (taken from What's New pages) below, but feel free to expand on or remove
from this list:
\begin{itemize}
    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
    \item APE 14 WCS
    \item A new SpectralCoord class for representing and transforming spectral quantities
    \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
\end{itemize}

@@

@@

Summary of Astroquery and engagement with archives. Cite astroquery paper.

@@

A prototype for the low-level analysis is ctapipe~\cite{ctapipe},
a python package developed to perform all the necessary tasks to from the raw data
of Cherenkov telescopes to the reconstructed event lists.

The high-level analysis (or CTA science tools) will be based on the astropy affiliated package
Gammapy~\cite{gammapy}.

@@

{\it Learn Astropy} is an umbrella term that acknowledges the broad educational
efforts made by the Astropy Project, which are led by the Learn Team.
The efforts focus on developing online content and workshops covering
astro-specific coding tasks in Python.
As described in \citet{astropy:2018}, There are four different types of
Learn Astropy content: \textit{tutorials}, consisting of Jupyter Notebook
lessons that are published in HTML format online; \textit{guides}, which are a
series of lessons providing a foundational resource for performing certain
type of astronomical analyses; \textit{examples}, which are snippets of code
that showcase a short task that can be performed with Astropy or an affiliated
package; and \textit{documentation}, which is contained within the code base.
This categorization drives content development, infrastructure choices, and the
appearance of the Learn Astropy website.
The Learn Team meets weekly to work on creating, expanding, improving these
educational resources.

\begin{itemize}

\item {\bf Learn Astropy website:} The Learn Team re-launched the website in
2021 with a new infrastructure platform, built around full-text search and
interactive filtering functionality, with the goal of making content
discoverable as the Learn Astropy content catalog expands.
This work has been supported in part by the Dunlap grant.
We have adopted Algolia, a search-as-a-service cloud platform, to store the
full-text and metadata records of Learn Astropy's content.
The new Learn Astropy website is a JavaScript (Gatsby/React) application that
uses the Algolia service to power its search and filtering user interface.
Our Python-based application, Learn Astropy Librarian, populates data into the
Algolia service. We tuned the Librarian around specific content formats (such as
Jupyter Notebook-based tutorial pages and Jupyter Book-based guides) to more
accurately index content and heuristically extract metadata.
A consequence of the new platform is that we now maintain and compile content
separately from the website application itself, enabling new content types.
Tutorials, which as Jupyter Notebook-based, are now compiled into their own
Learn Astropy sub-site using nbcollection.
Guides, which are Jupyter Book-based, are also deployed as separate websites
using GitHub pages. This architecture opens future possibilities of indexing
third-party content, hosted elsewhere, such as on institutional websites.

\item \textbf{Tutorials} Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.

\item \textbf{Guide} {\it ccdproc} Jupyter book has been completed.

\item {\bf Workshops:} the Project has been conducting workshops at winter
meetings of the American Astronomical Society since AAS 225 in January 2015.
Up to the start of the coronavirus pandemic, these were full-day in-person
workshops with as many as 90 participants and a dozen facilitators from the
project.
During the pandemic, these workshops were moved to an online format and split
into basic and advanced sessions.
Additionally, beginning with the AAS 238 online meeting, the workshops have been
expanded to Summer AAS meetings.
The Learn team finds that the workshop audience is best found as AAS meetings as
opposed to more general Python meetings.

The Astropy Project provided another mode of community engagement at AAS
Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in the
AAS Exhibit Hall.
Funding for the exhibit hall was provided alternately by NumFOCUS and later by
the Moore Foundation funding (\textbf{check this}).
The booth hosted a series of Q\&A special sessions during AAS 235 and webinars
during the virtual AAS 237 meetings, to provide the general astronomy community
information and access to experts on a variety of open source astronomical
tools.
\end{itemize}

@@

\begin{itemize}
\item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community. This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace. The Moore Foundation grant allows this Slack space to
be on a paid plan; additionally NumFOCUS has negotiated a special rate for
open-source projets. While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source. The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
a Discourse site which is more open than the Facebook group and more user-oriented
than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
developers are able to invoice as independent contractors the time they
spend helping users on these forums.
\end{itemize}
\secunfilled

>>>>

% Links to documentation: subpackages
\newcommand{\astropysubpkg}[1]{\href{http://docs.astropy.org/en/stable/#1/index.html}{\texttt{astropy.#1}}\xspace}
\newcommand{\astropyiosubpkg}[1]{\href{http://docs.astropy.org/en/stable/io/#1/index.html}{\texttt{astropy.io.#1}}\xspace}
\newcommand{\astropywcsaxes}{\href{http://docs.astropy.org/en/stable/visualization/wcsaxes/index.html}{\texttt{astropy.visualization.wcsaxes}}\xspace}
\newcommand{\astropycosmology}{\astropysubpkg{cosmology}}
\newcommand{\astropycosmologyunits}{\href{https://docs.astropy.org/en/stable/cosmology/units.html}{\texttt{astropy.cosmology.units}}}
\newcommand{\astropyunits}{\astropysubpkg{units}}
\newcommand{\astropycoordinates}{\astropysubpkg{coordinates}}
\newcommand{\astropyconstants}{\astropysubpkg{constants}}
\newcommand{\astropytime}{\astropysubpkg{time}}
\newcommand{\astropymodeling}{\astropysubpkg{modeling}}

% Links to documentation: classes
\newcommand{\astropyapi}[2]{\href{https://docs.astropy.org/en/stable/api/astropy.#1.html}{#2}}
\newcommand{\astropyapidoc}[2]{\astropyapi{#1}{\texttt{#2}\xspace}}

\newcommand{\astropySpectralCoord}{\astropyapidoc{coordinates.SpectralCoord}{SpectralCoord}}
\newcommand{\astropySkyCoord}{\astropyapidoc{coordinates.SkyCoord}{SkyCoord}}
\newcommand{\astropyICRS}{\astropyapidoc{coordinates.builtin_frames.ICRS}{ICRS}}
\newcommand{\astropyGalacticLSR}{\astropyapidoc{coordinates.builtin_frames.
GalacticLSR}{GalacticLSR}}
\newcommand{\astropyAltAz}{\astropyapidoc{coordinates.builtin_frames.AltAz}{AltAz}}
\newcommand{\astropyGalactocentric}{\astropyapidoc{coordinates.builtin_frames.Galactocentric}{Galactocentric}}
\newcommand{\astropyCosmology}{\astropyapidoc{cosmology.Cosmology}{Cosmology}}
\newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
\newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
\newcommand{\astropyQTable}{\astropyapidoc{table.QTable}{QTable}}
\newcommand{\astropyTime}{\astropyapidoc{time.Time}{Time}}
\newcommand{\astropyLeapSeconds}{\astropyapidoc{time.LeapSeconds}{LeapSeconds}}
\newcommand{\astropyDistribution}{\astropyapidoc{uncertainty.Distribution}{Distribution}}
\newcommand{\astropyUnit}{\astropyapidoc{units.Unit}{Unit}}
\newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
\newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}

% suggestion: shortcut reference to Paper II
\newcommand{\paperii}{\cite{astropy:2018}}

@@

The \astropypkg core package is mature and stable, in that many features have
been part of the core package for years with (purposefully) few changes to the
software interface.
This maturity and stability allows for the broader astronomy \python community
to rely upon the \astropypkg core package and build specialized packages within
the Astropy ecosystem.
Even many non-astronomical \python libraries have come to rely on \astropypkg.

The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citealt{astropy:2018}.
% feels like ↑ could be expanded to 2 sentences

% Maybe briefly outline the sectors of  
% Using \astropypkg for
% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
% Astropy has grown to interoperate, e.g. NumPy

Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.

% Summarize major pieces of development since the last paper (since v2.0). A few
% ideas (taken from What's New pages) below, but feel free to expand on or remove
% from this list:
% \begin{itemize}
%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
%     \item APE 14 WCS
%     \item A new SpectralCoord class for representing and transforming spectral quantities
%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
% \end{itemize}

\subsubsection*{New Features} \label{sec:core-features-new}

\begin{itemize}
  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
  \item Uncertainties and Distributions
  \item ERFA module is now it’s own thing; Obliquely tie in that units
    remains in astropy
  \item ? Pytest plugin ecosystem
  \item A new SpectralCoord class for representing and transforming spectral quantities
    Velocity in Coordinates
  \item WCS API : A shared Python interface for World Coordinate Systems
        Include APE reference
\end{itemize}

\subsubsection*{Interoperability} \label{sec:core-features-interoperability}

\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}

\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-performance}

\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-future}

@@

With support from the Moore Foundation, the \astropy Project was
allocated funding to support mentoring programs. In 2020 a call was
made to submit proposals for IDE (Inclusion, Diversity, and
Empowerment) initiatives for project-wide consideration on
Github. This process was deemed the most ``open'' because it allowed
for community wide feedback to focus and improve proposal
initiatives. Two programs were selected via this process, described
below.

\textbf{Outreachy:} \todo{No author assigned}{Describe the outreachy project}

\textbf{Women of Color Code (WoCCode)} is a peer-mentoring network for
coders from traditionally marginalized groups, most notably women of
color. Participants were invited to the WoCCode Slack space and
encouraged to attend monthly webinars to share skills in the context
of open source software libraries. Every other month, a guest speaker
was invited to talk about their career path and share a skill. Program
participants were solicited in the fall of 2020, yielding 73
applications from 17 countries (48\% from the United States, 37\% from
Africa, and 15\% from remaining continents). We selected participants
who we identified as having a high potential for contributing to open
source projects: intermediate to advanced programming skills with a
vocal interest in contributing. Of the thirty applicants invited to
the program, nineteen joined the community on Slack. Participants were
organized into cohorts based on interest and each cohort was assigned
one of three mentors that were also selected via open
application. Mentors acted as a general resource to participants, gave
one webinar, and organized a hack day. WoCCode also supported
registration of two participants to attend the American Astronomical
Society virtual summer 2021 meeting.

Participants rated the impact of the program as very high. In the
final webinar, participants reported a change in their perception of
coding in general, for example, accepting coding as something they can
do for fun. Participants reported generally feeling comfortable asking
questions and interacting with a community where ``[e]veryone's
thoughts are welcomed, no one is made to feel less important'' and
where one can enjoy ``warm interactions with likeminded people.''
WoCCode is continuing into 2022 by broadening participation in the
Slack space and publicly advertising the guest webinar events.

In addition to the supported mentoring initiatives, the \astropy
project as a whole has taken steps to examine representation of
marginalized groups within the project and search for avenues of
improvement.

% Much of this text was adapted from Robel Geda's report from NSBP and
% SACNAS. Please include him on the author list.
\textbf{\astropy representation at national diversity conferences:}
Several members of the \astropy community attended virtual conferences
of the National Society of Black Physicists (NSBP) and SACNAS
(focusing on Hispanic and Native American scientists across all STEM
fields). \astropy representatives noted that one underlying topic came
up multiple times. A major barrier for persons who come from
underrepresented communities is the lack of resources and expertise
that are necessary to train students. Writing PEP 8 compliant software
and understanding the Git and Github workflow is not part of the
standard Physics and Astrophysics curriculum. Additionally, the larger
Astronomnical community is still in the process of transitioning
towards Python based tools, making it difficult for students not
currently under advisement by some one with extensive Python expertise
to get involved with the open source Python community. Projects such
as Learn Astropy could have a profound impact by empowering
underrepresented groups because it provides a free, searchable, and
accessible introduction to Python tools for astrophysical
research. Representatives also noted that offering training and
teaching materials to PIs of Research Experience for Undergraduate
(REU) programs throughout the United States could be helpful for
encouraging advisors to teach students how to use Astropy, create
their own libraries, and use version control on open source platforms
like Github. Such materials could also be offered as a workshop at
national diversity conferences themselves, as SACNAS solicits special
session proposals each year.

\textbf{NumFOCUS ``Contributor Diversification \& Retention'' (CDR)
  initiative:} \todo{No author assigned}{Describe more about this.}

@@

Astroquery \cite{Ginsburg2019} is the astropy-coordinated package for interacting with online archives of astronomical and related data. It contains over 50 modules for querying astronomical databases, large and small. In particular, since \paperii, significant contributions to \texttt{astroquery} have come from several of the major archives, including the European Space Agency (ESA), the Mikulski Archive for Space Telescopes (MAST) at the Space Telescope Science Institute (STScI), the Infrared Science Archive at the NASA Infrared Processing and Analysis Center (IRSA at NASA IPAC), the Canadian Astronomical Data Center (CADC), and the Atacama Large Millimeter/Submillimeter Array archive (ALMA). These contributions represent a formal acknowledgement of the utility of a centralized tool suite for archive interaction.

The widespread usage of \texttt{astroquery} is apparent from the range of keywords represented in the journal articles which cite \cite{Ginsburg2019}; everything from astroids to galaxies is represented. Additionally Astroquery functionality has been built into several special purpose Python packages, such as \texttt{LightKurve} \cite{LightKurve} and SORA \cite{SORA}]. More than 4,000 repositories on GitHub make use of \texttt{astroquery} in some way, an \texttt{astroquery} features in a large number of tutorials on various facets of astronomical analysis.

Many of the existing and newly-contributed tools rely on Virtual Observatory (VO) tools.
These use the underlying package \texttt{pyvo}, which has also recently become an astropy-coordinated package. While many or all of the functions provided in \texttt{astroquery} can be achieved through direct use of VO tools implemented in \texttt{pyvo}, the \texttt{astroquery} interfaces more closely resemble the web interfaces more familiar to users.
% note to coauthors: we need a section on pyvo

@@

A prototype for the low-level analysis is \texttt{ctapipe} \citep{ctapipe},
a python package developed to perform all the necessary tasks to from the raw data
of Cherenkov telescopes to the reconstructed event lists.

The high-level analysis (or CTA science tools) will be based on the astropy affiliated package
Gammapy~\citep{gammapy}.

@@

\textit{Learn Astropy} is an umbrella term that acknowledges the broad
educational efforts made by the Astropy Project, which are led by the Learn
Team.
The efforts focus on developing online content and workshops covering
astronomy-specific coding tasks in Python.
As introduced in \citet{astropy:2018}, there are four different types of
Learn Astropy content: \textit{tutorials}, consisting of Jupyter Notebook
lessons that are published in HTML format online; \textit{guides}, which are a
series of lessons providing a foundational resource for performing certain
type of astronomical analyses; \textit{examples}, which are snippets of code
that showcase a short task that can be performed with Astropy or an affiliated
package; and \textit{documentation}, which is contained within the code base.
This categorization drives content development, infrastructure choices, and the
appearance of the Learn Astropy website.
The Learn Team meets weekly to work on creating, expanding, improving these
educational resources.

\begin{description}

\item[Learn Astropy website] The Learn Team re-launched the main website and
search interface for Learn Astropy in 2021 with a new infrastructure platform,
built around full-text search and interactive filtering functionality, with the
goal of making content discoverable as the Learn Astropy content catalog
expands.
This work has been supported in part by a grant from the Dunlap Institute.
We have adopted Algolia, a search-as-a-service cloud platform, to store the
full-text and metadata records of Learn Astropy's content.
The new Learn Astropy website is a JavaScript (Gatsby/React) application that
uses the Algolia service to power its search and filtering user interface.
Our Python-based application, Learn Astropy Librarian, populates data into the
Algolia service. We tuned the Librarian around specific content formats (such as
Jupyter Notebook-based tutorial pages and Jupyter Book-based guides) to more
accurately index content and heuristically extract metadata.
A consequence of the new platform is that we now maintain and compile content
separately from the website application itself, enabling new content types.
Tutorials, which are written as Jupyter Notebooks, are now compiled into their
own Learn Astropy sub-site using
\href{https://github.com/astropy/nbcollection}{\texttt{nbcollection}}.
Guides, which utilize the \href{https://jupyterbook.org/}{Jupyter Book} build
infrastructure, are also deployed as separate websites using GitHub pages.
This architecture opens future possibilities of indexing third-party content,
hosted elsewhere, such as on institutional websites.

\item[Tutorials] Currently, we have \# of tutorials, spanning a wide range of astronomical topics and common tasks.

\item[Guides] {\it ccdproc} Jupyter book has been completed.

\item[Workshops] The Project has been conducting workshops at winter
meetings of the American Astronomical Society since AAS 225 in January 2015.
Up to the start of the coronavirus pandemic, these were full-day in-person
workshops with as many as 90 participants and a dozen facilitators from the
project.
During the pandemic, these workshops were moved to an online format and split
into basic and advanced sessions.
Additionally, beginning with the AAS 238 online meeting, the workshops have been
expanded to Summer AAS meetings.
The Learn team finds that the workshop audience is best found as AAS meetings as
opposed to more general Python meetings.

The Astropy Project provided another mode of community engagement at AAS
Meetings 235 and 237 by organizing a NumFOCUS Sponsored Projects booth in the
AAS Exhibit Hall.
Funding for the exhibit hall was provided alternately by NumFOCUS and later by
the Moore Foundation funding (\textbf{check this}).
The booth hosted a series of Q\&A special sessions during AAS 235 and webinars
during the virtual AAS 237 meetings, to provide the general astronomy community
information and access to experts on a variety of open source astronomical
tools.
\end{description}

@@

\begin{itemize}
\item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community. This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace. The Moore Foundation grant allows this Slack space to
be on a paid plan; additionally NumFOCUS has negotiated a special rate for
open-source projects. While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source. The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
a Discourse site which is more open than the Facebook group and more user-oriented
than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
developers are able to invoice as independent contractors the time they
spend helping users on these forums.
\end{itemize}
